Highly Available Transactions: Virtues and Limitations\iftechreport\\(Extended Version)\fi

### 2013-10-04 writeup/evaluation.tex
more edits

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Here, we verify that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties (over WAN, 10 to 100 times higher than
their HAT counterparts) compared to HAT-compliant algorithms, which
scale linearly.
x
\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based partitioned) key-value backed by
LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (hereafter, \texttt{eventual};
last-writer-wins RU with standard all-to-all anti-entropy between
replicas) and the efficient HAT MAV algorithm as sketched in
Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). We support non-HAT
operation whereby all operations for a given key are routed to a
(randomly) designated \texttt{master} replica for each key
(guaranteeing single-key linearizability, as in Gilbert and Lynch's
CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
``read latest'' operation; hereafter, \texttt{master}) as well as
distributed two-phase locking. Servers are durable: they synchronously
write to LevelDB before responding to client requests, while new
writes in MAV are synchronously flushed to a disk-resident write-ahead
log.

>>>>

To demonstrate the performance implications of HAT guarantees in a
real-world environment, we implemented a HAT database prototype. We
verify that, as Section~\ref{sec:latency}'s measurements suggested,
``strongly consistent'' algorithms incur substantial latency penalties
(over WAN, 10 to 100 times higher than their HAT counterparts)
compared to HAT-compliant algorithms, which scale linearly. Our goal
is \textit{not} a complete performance analysis of HAT semantics but
instead a proof-of-concept demonstration of a small subset of the HAT
space on real-world infrastructure.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based partitioned) key-value backed by
LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (hereafter, \texttt{eventual};
last-writer-wins RU with standard all-to-all anti-entropy between
replicas) and the efficient HAT MAV algorithm as sketched in
Section~\ref{sec:ta}. (hereafter, \texttt{MAV}). We support non-HAT
operation whereby all operations for a given key are routed to a
(randomly) designated \texttt{master} replica for each key
(guaranteeing single-key linearizability, as in Gilbert and Lynch's
CAP Theorem proof~\cite{gilbert-cap} and in PNUTS~\cite{pnuts}'s
``read latest'' operation; hereafter, \texttt{master}) as well as
distributed two-phase locking. Servers are durable: they synchronously
write to LevelDB before responding to client requests, while new
writes in MAV are synchronously flushed to a disk-resident write-ahead
log.

### 2013-10-04 writeup/hat-analysis.tex
more edits

Several different properties have been labeled \textbf{Repeatable
  Read} isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures the spirit of the term: if a
transaction reads the same data more than once, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation}, and, if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
the stronger property of \textbf{Predicate Cut-Isolation}. In the
example below, under both levels of cut isolation, $T_3$ must read
$a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having transactions cache their reads at the client such that the
values that they read for each item never changes unless they
overwrite it themselves. The cache can be cleared at the end of each
transaction and can alternatively be accomplished on (sticky) replicas
via multi-versioning. Predicate Cut Isolation is also achievable in
HAT systems via similar caching or multi-versioning that track entire
logical ranges of predicates in addition to item-based reads.

@@

In light the of current practice of deploying weak isolation levels
(Section~\ref{sec:modernacid}), it is perhaps surprising that so many
weak isolation levels are achievable as HATs. Indeed, isolation levels
such as Read Committed expose and are defined in terms of end-user
anomalies that could not arise during serializable execution. However,
the prevalence of these models suggests that, in many cases,
applications can tolerate these their associated anomalies. Given our
HAT-compliance results, this in turn hints that--despite
idiosyncrasies relating to concurrent updates and data recency--highly
available database systems can provide sufficiently strong semantics
for many applications. Indeed, HAT databases may expose more anomalies
than a single-site database operating under weak isolation
(particularly during network partitions). However, for a fixed
isolation level (which, in practice, can vary across databases and may
differ from implementation-agnostic definitions in the literature),
users of single-site database are subject to the same (worst-case)
application-level anomalies as a HAT implementation. The necessary
(indefinite) visibility penalties (i.e., the right side of
Figure~\ref{fig:hat-order}) and lack of support for preventing
concurrent updates (via the upper left half of
Figure~\ref{fig:hat-order}) mean HATs are \textit{not} well-suited for
all applications (see Section~\ref{sec:evaluation}): these limitations
are fundamental. However, common practices such as ad-hoc, user-level
compensation and per-statement isolation ``upgrades'' (e.g.,
\texttt{SELECT FOR UPDATE} under weak isolation)--commonly used to
augment weak isolation--are also applicable in HAT systems (although
they may in turn compromise availability).

>>>>

Several different properties have been labeled \textbf{Repeatable
  Read} isolation. As we will show in
Section~\ref{sec:unachievable-acid}, some of these are not achievable
in a HAT system. However, the ANSI standardized
implementation-agnostic definition~\cite{ansi-sql} \textit{is}
achievable and directly captures the spirit of the term: if a
transaction reads the same data more than once, it sees the same value each
time (preventing ``Fuzzy Read,'' or $P2$). In this paper, to
disambiguate between other definitions of ``Repeatable Read,'' we will
call this property ``cut isolation,'' since each transaction reads
from a non-changing cut, or snapshot, over the data items. If this
property holds over reads from discrete data items, we call it
\textbf{Item Cut Isolation}, and, if we also expect a cut over
predicate-based reads (e.g., \texttt{SELECT WHERE}; preventing
Phantoms~\cite{gray-isolation}, or Berenson et al.'s $P3/A3$), we have
the stronger property of \textbf{Predicate Cut-Isolation}. In the
example below, under both levels of cut isolation, $T_3$ must read
$a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
It is possible to satisfy Item Cut Isolation with high availability by
having transactions store a copy of any read data at the client such
that the values that they read for each item never changes unless they
overwrite it themselves. These stored values can be discarded at the
end of each transaction and can alternatively be accomplished on
(sticky) replicas via multi-versioning. Predicate Cut Isolation is
also achievable in HAT systems via similar caching middleware or
multi-versioning that track entire logical ranges of predicates in
addition to item-based reads.

@@

In light the of current practice of deploying weak isolation levels
(Section~\ref{sec:modernacid}), it is perhaps surprising that so many
weak isolation levels are achievable as HATs. Indeed, isolation levels
such as Read Committed expose and are defined in terms of end-user
anomalies that could not arise during serializable execution. However,
the prevalence of these models suggests that, in many cases,
applications can tolerate these their associated anomalies. Given our
HAT-compliance results, this in turn hints that--despite
idiosyncrasies relating to concurrent updates and data recency--highly
available database systems can provide sufficiently strong semantics
for many applications. Indeed, HAT databases may expose more anomalies
than a single-site database operating under weak isolation
(particularly during network partitions). However, for a fixed
isolation level (which, in practice, can vary across databases and may
differ from implementation-agnostic definitions in the literature),
users of single-site database are subject to the same (worst-case)
application-level anomalies as a HAT implementation. The necessary
(indefinite) visibility penalties (i.e., the right side of
Figure~\ref{fig:hatcompared}) and lack of support for preventing
concurrent updates (via the upper left half of
Figure~\ref{fig:hatcompared}) mean HATs are \textit{not} well-suited
for all applications (see Section~\ref{sec:evaluation}): these
limitations are fundamental. However, common practices such as ad-hoc,
user-level compensation and per-statement isolation ``upgrades''
(e.g., \texttt{SELECT FOR UPDATE} under weak isolation)--commonly used
to augment weak isolation--are also applicable in HAT systems
(although they may in turn compromise availability).

### 2013-10-04 writeup/hat-appendix.tex
more edits

\begin{definition}[Read Atomic (RA)]
A system that provides Read Atomic isolation prohibits phenomenon
OTV in addition to providing Read Committed isolation.
\end{definition}

@@

\section{Read Atomic Isolation}

In this section, we provide additional information on our two-phase
algorithm for providing Read Atomic isolation.

@@

\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: r_x(1)~r_y(a)
\end{align*}
For RA, the system must ensure that $a=1$. Consider a system with one
replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
(whereby the last $4$ digits are reserved for client ID to ensure
uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
their respective writes into their individual \texttt{pending} sets
and send acknowledgment for transaction timestamped $10001$ to the
other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
its \texttt{required} vector so that $\texttt{required}=\{x:10001,
y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
$ts=10001$ and $R_y$ will not find a suitable write
in \texttt{good}. $R_y$ will instead return $y=1$
from \texttt{pending}.

\vspace{1em}\noindent\textbf{Overheads.} RA requires overhead both on client and server. First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read. Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction. (Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network. Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}. If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value. The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.

@@

\vspace{1em}\noindent\textbf{Algorithmic improvements.} We are actively investigating more efficient alternatives to the algorithm presented here. In particular, a lower-overhead (but still highly available) implementation that used less metadata would be appealing. A system using server-side vector clocks is feasible but does not necessarily reduce overhead because the number of items written per transaction is likely smaller than the number of servers in a system. Similarly, we have observed an algorithm trade-off between the size of $ts_{required}$ and write metadata. While it is certainly possible to reduce overheads by reducing availability~\cite{eiger}, we believe high availability is a worthwhile goal for future exploration.

>>>>

\begin{definition}[Monotonic Atomic View (MAV)]
A system that provides Monotonic Atomic View isolation prohibits
phenomenon OTV in addition to providing Read Committed isolation.
\end{definition}

@@

\section{Monotonic Atomic View Isolation}

In this section, we provide additional information on our two-phase
algorithm for providing Monotonic Atomic View isolation. This section
is effectively a work-in-progress; we are actively investgating
alternative algorithms (particularly with respect to metadata
overheads) and stronger (but still HAT) semantics that complement the
techniques presented here. The interested reader (or would-be systems
implementer) should contact Peter Bailis for more details.

@@

\vspace{1em}\noindent\textbf{Example.} To illustrate this algorithm consider an execution of the following transactions:
\begin{align*}
\small\vspace{-1em}
T_1 &: w_x(1)~w_y(1)
\\T_2 &: r_x(1)~r_y(a)
\end{align*}
For MAV, the system must ensure that $a=1$. Consider a system with one
replica for each of $x$ and $y$ (denoted $R_x$ and $R_y$). Say client
$c_1$ executes $T_1$ and commits. The client chooses timestamp $10001$
(whereby the last $4$ digits are reserved for client ID to ensure
uniqueness) and sends values to $w_x(1)$ to $R_x$ and $R_y$ with
$tx\_keys = \{x,y\}$ and $timestamp=10001$. $R_x$ and $R_y$ place
their respective writes into their individual \texttt{pending} sets
and send acknowledgment for transaction timestamped $10001$ to the
other. Consider the case where $r_x$ has seen $r_y$'s acknowledgment
but not vice versa: $w_x(1)$ will be in \texttt{good} on $R_x$
but \texttt{pending} on $R_y$. Now, if another client $c_2$ executes
$T_2$ and reads $x$ from $R_x$, it will read $x=1$. $c_2$ will update
its \texttt{required} vector so that $\texttt{required}=\{x:10001,
y:10001\}$. When $c_2$ reads $y$ from $R_y$, it will specify
$ts=10001$ and $R_y$ will not find a suitable write
in \texttt{good}. $R_y$ will instead return $y=1$
from \texttt{pending}.

\vspace{1em}\noindent\textbf{Overheads.} MAV requires overhead both on client and server. First, clients must store a vector (\texttt{required}) for the duration of a transaction in order to determine what values to read. Second, each write contains metadata ($tx\_keys$) linear in the number of writes in the transaction. (Note that we could have stored a different timestamp with every write, but a single timestamp suffices and is more efficient) These overheads are modest for short transaction lengths but they increase value sizes on disk, in memory, and over the network. Third, on the server side, each client \textsc{put} incurs two back-ends \texttt{put}s: one into \texttt{pending} and \texttt{good}. If keeping a single value for \texttt{good}, the second \texttt{put} into \texttt{good} (as is the case with normal, single-valued eventually consistent operation) is really a \texttt{get} then a possible \texttt{put}: the server first checks if a higher-timestamped value has been written and, if not, it overwrites the current value. The \texttt{put} into \texttt{pending} is simpler: if message delivery is exactly-once, then \texttt{put}s can be made unconditionally.

@@

### 2013-10-04 writeup/model.tex
more edits

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical (sets of) replica(s) across subsequent
operations. As we will discuss in Section~\ref{sec:hats}, clients can
ensure continuity between operations (e.g., reading their prior
updates to a data item) by maintaining affinity or ``stickiness'' with
a server or set of servers~\cite{vogels-defs}. In a fully-replicated
system, where all servers are replicas for all data items, stickiness
is simple: a client can maintain stickiness by contacting the same
server for each of its requests. However, to stay ``sticky'' in a
partially-replicated system, where servers are replicas for subsets of
the set of data items (which we consider in this paper), a client must
maintain stickiness with a single \textit{logical} server that may
consist of multiple physical servers. We say that a system provides
\textbf{sticky availability} if, whenever a client's operation on an
item is executed against a logical server that has observed all of its
prior operations, it eventually receives a response, even in the
presence of indefinitely long partitions. A client may choose to
become sticky available by acting as a server itself; for example, a
client might cache its reads and writes~\cite{bolton,
  sessionguarantees, swift}. Any guarantee achievable in a highly
available system is achievable in a sticky high availability system
but not vice-versa.

>>>>

In addition to high availability, which allows operations on any
replica, distributed algorithms often assume a model in which clients
always contact the same logical replica(s) across subsequent
operations, whereby each of the client's prior operations (but not
necessarily other clients' operations) are reflected in the database
state that they observe. As we will discuss in Section~\ref{sec:hats},
clients can ensure continuity between operations (e.g., reading their
prior updates to a data item) by maintaining affinity or
``stickiness'' with a server or set of servers~\cite{vogels-defs}. In
a fully-replicated system, where all servers are replicas for all data
items, stickiness is simple: a client can maintain stickiness by
contacting the same server for each of its requests. However, to stay
``sticky'' in a partially-replicated system, where servers are
replicas for subsets of the set of data items (which we consider in
this paper), a client must maintain stickiness with a single
\textit{logical} copy of the database, which may consist of multiple
physical servers. We say that a system provides \textbf{sticky
  availability} if, whenever a client's transactions is executed
against a copy of database state that reflects all of the client's prior
transactions, it eventually receives a response, even in the presence
of indefinitely long partitions. A client may choose to become sticky
available by acting as a server itself; for example, a client might
cache its reads and writes~\cite{bolton, sessionguarantees,
  swift}. Any guarantee achievable in a highly available system is
achievable in a sticky high availability system but not vice-versa.
